---
title: "RF"
author: "David Oloyede & Khalyl Smith"
date: "4/29/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, message = FALSE, warning=FALSE}
library(tidyverse)
base_stroke<- read.csv("~/stroke-data.csv")
head(base_stroke)

base_stroke$Indice<-c(1:nrow(base_stroke))
base_stroke<-base_stroke[,c(13,1:12)]
base_stroke$ever_married<-as_factor(base_stroke$ever_married)
base_stroke$work_type<-as_factor(base_stroke$work_type)
base_stroke$Residence_type<-as_factor(base_stroke$Residence_type)
base_stroke$smoking_status<-as_factor(base_stroke$smoking_status)

base_stroke$stroke<-as_factor(base_stroke$stroke)

base_stroke$bmi<-as.numeric(base_stroke$bmi)

head(base_stroke,4)

qtd_NAs<-c()

for(i in 1 : length(base_stroke)){
    
    qtd_NAs[i]<-sum(is.na(base_stroke[,i]))
}
qtd_NAs

media_bmi<-base_stroke %>% group_by(gender) %>% summarize(media_Bmi=mean(bmi,na.rm = TRUE),n=n())

for(i in 1: nrow(base_stroke)){
    if(is.na(base_stroke$bmi[i])){
        if(base_stroke$gender[i] == "Male"){
            base_stroke$bmi[i]<-media_bmi$media_Bmi[1]
        }
        if(base_stroke$gender[i] == "Female"){
            base_stroke$bmi[i]<-media_bmi$media_Bmi[2]
        }
        if(base_stroke$gender[i] == "Other"){
            base_stroke$bmi[i]<-media_bmi$media_Bmi[3]
        }
    }
}
levels(base_stroke$stroke)<-c("No stroke","Had stroke")
```

Since it is a classification task and we are predicting the reponse for stroke, we are using random forests. An advantage of random forests is that it allows for a reduction in variance, and thus lower test error, compared to both single decision trees and tree bagging as well as handling categorical, continuous, and non-linear parameters efficiently without need for scaling.First we must split data.
Here we are splitting the database and excluding the first two columns.The data is split into training set and testing set with an 70-30 split.   
```{r, message = FALSE, warning=FALSE}
base_stroke<-base_stroke[,-c(1:2)]
library(caTools)
set.seed(10)
div<-sample.split(Y = base_stroke$stroke,SplitRatio = 0.70)
base_training<-subset(base_stroke,subset = div == TRUE)
base_test<-subset(base_stroke,subset = div == FALSE)

```

For the random forest model, we'll use `stroke` as our response variable and all other variables (excluding for `id`) as our predictors. This is the formula we will use for the model:
  
stroke ∼ gender + age + hypertension + heart_disease + ever_married+
work_type + Residence_type + avg_glucose_level + bmi + smoking_status   

A longer training period and greater complexity in interpretation as a result of multiple trees is the cost of having all the advantages of the model explained previously. This model will consist of 20 trees. 
```{r, message = FALSE, warning=FALSE}
library(randomForest)
```

```{r}
set.seed(1)
model_RF<-randomForest(formula = stroke ~.,data = base_training,ntree =20)
model_RF

```
The Random Forests had an error rate of 5.68% 


```{r}
prediction<-predict(model_RF,newdata = base_test[,-11])

confusionMatrix<-table(base_test$stroke,prediction)
confusionMatrix

```

```{r}
accuracy<-(confusionMatrix[1] + confusionMatrix[4])/ sum(confusionMatrix)
accuracy

```
With an accuracy rate of 95% we can see that the model is not very good however of predicting risk of having a stroke with a bias of not having a stroke. Even though random forests model are not as computationally expensive as alot of others and can reduce high variance. 





